---
title: "Getting started with ADEPT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ADEPT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7,
  fig.height = 4.2
)
```

```{r}
library(ADEPT)
```

## What ADEPT does

LA-ICP-MS depth profiling acquires a time series of U-Pb isotope ratios as the
laser drills into a zircon. Downhole fractionation and real age differences
both make the apparent age drift, so choosing "the age" means deciding *where*
in the profile the signal is a plateau.

ADEPT automates that decision:

1. discard outlying ages,
2. smooth the series with LOESS and standardise it,
3. segment it with PELT changepoint detection,
4. filter the segments to a set of plateaus consistent with the expected age
   trend,
5. report an age and a full uncertainty budget for each plateau.

The method is described in Wang et al. (2026), *JGR: Solid Earth*.

## Input data

Three formats are accepted, distinguished by column names. Column names must
match exactly, including case.

| Format | Required columns |
|---|---|
| 1. Direct ages | `Age68`, `Age75`, `Age76` |
| 2. Raw isotope counts | `Pb206`, `Pb207`, `U238` |
| 3. Isotopic ratios | `Pb206_U238`, `Pb207_U235`, `Pb207_Pb206` |

All three need `Analysis` (sample name) and `Time` (seconds). Any extra numeric
columns are summarised per plateau and carried into the output. Formats 2 and 3
are converted with the decay constants and the ²³⁸U/²³⁵U ratio used by
IsoplotR, so the ages match what you would get there.

## A first run

```{r}
input <- system.file("extdata", "Input(1sample).xlsx", package = "ADEPT")
result <- adept(input, output_path = NA, make_plots = FALSE, verbose = FALSE)
```

`output_path = NA` keeps everything in memory. Normally `adept()` also writes a
two-sheet Excel workbook (all segments, and a summary of the confirmed
plateaus) and a PDF per zircon.

```{r}
result$summary[, c("Final serial number", "Integration time",
                   "Final age (Ma)",
                   "Final total uncertainty incl. decay (Ma)",
                   "Confirmed plateaus")]
```

## Reading the depth profile

`keep_profiles = TRUE` returns the smoothed series and the plateau table per
zircon, which is what you need for a figure.

```{r fig.height = 3.6}
r <- adept(input, output_path = NA, make_plots = TRUE, keep_profiles = TRUE,
           verbose = FALSE)
# the objects in $plots carry a print() method that draws the profile
r$plots[[1]]
```

Orange dots are the raw ²⁰⁶Pb/²³⁸U ages, the black line is the LOESS trend, and
each red bar is a confirmed plateau with its uncertainty band.

## The uncertainty budget

This is the part worth reading carefully. Each plateau comes with a
decomposition, because "± 0.7 Ma" on its own does not tell a reader whether
that number was measured or assumed.

```{r}
keep <- !is.na(result$summary[["Final age (Ma)"]])
result$summary[keep, c("Final age (Ma)",
                       "Random uncertainty (Ma)",
                       "Systematic uncertainty (Ma)",
                       "Decay constant uncertainty (Ma)",
                       "Relative uncertainty (%)")]
```

- **Random** is what you compare ages with inside a session: the scatter of the
  points inside the plateau, plus the reproducibility of the primary reference
  material (`calibration_uncertainty`, 3 % by default).
- **Systematic** is the decay-constant contribution. It is fully correlated
  between samples, so it does not cancel when you average ages.

On this example the reference-material term dominates: the measured
within-plateau scatter contributes only a few percent of the total. **The
default 3 % is an assumption, not a measurement** — replace it with your
laboratory's long-term reproducibility before quoting the age, and say where
the number came from.

```{r}
smaller <- adept(input, calibration_uncertainty = 0.015,
                 output_path = NA, make_plots = FALSE, verbose = FALSE)
data.frame(assumed = c("3 %", "1.5 %"),
           random = c(result$summary[["Random uncertainty (Ma)"]][1],
                      smaller$summary[["Random uncertainty (Ma)"]][1]),
           age = result$summary[["Final age (Ma)"]][1])
```

The age itself does not move; only the uncertainty does.

## Is the plateau internally consistent? (MSWD)

Everything above answers "how precisely do we know this age". MSWD answers a
different question: **do the points inside the plateau actually agree with one
another?** A plateau can look precise and still be a mixture of two
populations.

To ask that question, the input needs a per-point 1-sigma column. Below is a
synthetic profile containing two real ages, 100 and 130 Ma, whose points
scatter by 3 Ma - but which claims an uncertainty of only 1.5 Ma:

```{r}
set.seed(1)
age <- c(rnorm(133, 100, 3), rnorm(67, 130, 3))
prof <- data.frame(
  Analysis = "demo",
  Time     = seq(0, 60, length.out = 200),
  Age68    = age,
  Age75    = age + rnorm(200, 0, 0.5),
  Age76    = age + rnorm(200, 0, 0.5),
  Age68_1s = 1.5
)
csv <- file.path(tempdir(), "demo.csv")
write.csv(prof, csv, row.names = FALSE)

w <- adept(csv, output_path = NA, make_plots = FALSE, verbose = FALSE)
w$summary[, c("Final age (Ma)", "Random uncertainty (Ma)",
              "MSWD", "MSWD probability", "Points with uncertainty")]
```

The ages come out right, and MSWD tells you not to trust the uncertainties: a
value near 4 is what you expect when the real scatter is twice the quoted
sigma. Read it like this:

- **MSWD around 1** - the spread matches the errors you supplied, and the
  plateau can be read as a single age.
- **MSWD much above 1** - the points scatter more than their errors allow.
  Either the plateau really spans more than one age, or the sigmas are too
  small. The usual cause of the latter is counting statistics quoted without
  the external reproducibility of the reference material.
- **MSWD much below 1** - the sigmas are too large.

`MSWD probability` is the tail probability under the consistent case; below
about 0.05 is the conventional flag. Here it is essentially zero, which is the
correct answer for this dataset: the sigmas really are too small.

Note also that with a sigma column the plateau age changes character. It
becomes the inverse-variance weighted mean of the measured ages, so a point
measured to 0.5 Ma counts a hundred times as much as one measured to 5 Ma.
That is why these ages differ slightly from the same profile run without
sigmas.

## Choosing the parameters, and knowing how much they matter

Two parameters control how many plateaus you get:

- `variance_threshold` (default 0.1192) — how internally consistent a segment
  has to be;
- `min_plateau_resolution` (default 5 s) — how long a plateau has to last.

Both defaults sit close to a cliff. `adept_sensitivity()` maps out the
neighbourhood so you can report it rather than discover it in review.

```{r}
s <- adept_sensitivity(input,
                       variance_threshold = c(0.08, 0.1192, 0.25),
                       min_plateau_resolution = c(5, 10),
                       filter_direction = "Forward",
                       verbose = FALSE)
s[, c("variance_threshold", "min_plateau_resolution", "n_plateaus", "ages")]
```

Put a table like this in your supplementary material. If the plateau count
changes from 0 to 4 across a plausible range of a parameter, the reader
deserves to know.

`filter_direction` is the third choice: `"Forward"` keeps ages that increase
downhole, `"Reverse"` keeps ages that decrease. It is not a cosmetic switch —
on this example Forward returns three plateaus and Reverse returns one — so it
has to be justified from the CL images, not from which answer you prefer.

## Pre-processing options

Two pipelines are available:

- `preprocess = "arima_loess"` (default) — screen outliers with an ARIMA
  residual test, then fit an ordinary LOESS.
- `preprocess = "robust_loess"` — skip the screening and fit
  `loess(family = "symmetric")`, which downweights outliers instead of deleting
  them. This removes the least stable step, since it has no model-order search
  to go wrong.

```{r}
rb <- adept(input, preprocess = "robust_loess", output_path = NA,
            make_plots = FALSE, verbose = FALSE)
data.frame(preprocess = c("arima_loess", "robust_loess"),
           plateaus = c(nrow(result$summary), nrow(rb$summary)),
           ages = c(paste(signif(result$summary[["Final age (Ma)"]], 6),
                          collapse = "; "),
                    paste(signif(rb$summary[["Final age (Ma)"]], 6),
                          collapse = "; ")))
```

Run both and report the comparison. If the geological conclusion changes
between them, the conclusion was never robust.

There is a third option, `smooth = "none"`, which turns smoothing off
altogether and segments the profile as measured. It exists for data that have
already been corrected for down-hole fractionation before they reach ADEPT -
say an F(tau) correction applied by an external reduction. Such a profile is
already flat inside a domain, and fitting LOESS to it a second time would
round off the real domain boundaries.

```{r}
ns <- adept(input, smooth = "none", output_path = NA, make_plots = FALSE,
            verbose = FALSE)
data.frame(setting = c("smooth = \"loess\" (default)", "smooth = \"none\""),
           plateaus = c(nrow(result$summary), nrow(ns$summary)),
           first_age = c(result$summary[["Final age (Ma)"]][1],
                         ns$summary[["Final age (Ma)"]][1]))
```

The ages move by a small amount rather than changing the answer. What changes
is that the segmentation follows the profile you gave it, instead of a
smoothed version of it. On uncorrected data, leave the default.

## Bayesian posterior (optional)

`mcmc = TRUE` adds a piecewise-linear changepoint posterior via the `mcp`
package (which needs JAGS). It is entirely optional: without `mcp` installed
the MCMC columns are `NA` and the rest of the pipeline runs normally.

## Before you interpret the ages

A plateau is the best piecewise-constant description of the data under the
parameters you chose. That is not the same as a geological event.

For zircons that have experienced multi-stage disturbance — uranium ore,
hydrothermally altered grains — the U-Pb system may not have behaved as a
closed system in any interval, and no amount of segmentation will recover a
meaningful age. Always check the plateaus against CL/BSE imaging and the trace
element profiles (`Th_U_Mean`, `Pb_Mean` and friends are carried through for
exactly this).

Where the package can help it does: ages outside 0–4568 Ma, non-monotonic
`Time`, and a `Time` axis that looks like milliseconds are all reported before
the model is fitted.

## References

- Wang, Y. et al. (2026). Automated Depth Profiling Technique. *JGR: Solid
  Earth*, 131, e2025JB033324.
- Jaffey, A. H. et al. (1971). Precision measurement of half-lives and specific
  activities of ²³⁵U and ²³⁸U. *Physical Review C*, 4, 1889.
- Hiess, J. et al. (2012). ²³⁸U/²³⁵U systematics in terrestrial uranium-bearing
  minerals. *Science*, 335, 1610.
- Killick, R., Fearnhead, P. & Eckley, I. A. (2012). Optimal detection of
  changepoints with a linear computational cost. *JASA*, 107, 1590.
