---
title: "Getting started with ADEPT"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ADEPT}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 7,
  fig.height = 4.2
)
```

```{r}
library(ADEPT)
```

## What ADEPT does

LA-ICP-MS depth profiling acquires a time series of U-Pb isotope ratios as the
laser drills into a zircon. Downhole fractionation and real age differences
both make the apparent age drift, so choosing "the age" means deciding *where*
in the profile the signal is a plateau.

ADEPT automates that decision:

1. discard outlying ages,
2. smooth the series with LOESS and standardise it,
3. segment it with PELT changepoint detection,
4. filter the segments to a set of plateaus consistent with the expected age
   trend,
5. report an age and a full uncertainty budget for each plateau.

The method is described in Wang et al. (2026), *JGR: Solid Earth*.

## Input data

Three formats are accepted, distinguished by column names. Column names must
match exactly, including case.

| Format | Required columns |
|---|---|
| 1. Direct ages | `Age68`, `Age75`, `Age76` |
| 2. Raw isotope counts | `Pb206`, `Pb207`, `U238` |
| 3. Isotopic ratios | `Pb206_U238`, `Pb207_U235`, `Pb207_Pb206` |

All three need `Analysis` (sample name) and `Time` (seconds). Any extra numeric
columns are summarised per plateau and carried into the output. Formats 2 and 3
are converted with the decay constants and the ²³⁸U/²³⁵U ratio used by
IsoplotR, so the ages match what you would get there.

## A first run

```{r}
input <- system.file("extdata", "Input(1sample).xlsx", package = "ADEPT")
result <- adept(input, output_path = NA, make_plots = FALSE, verbose = FALSE)
```

`output_path = NA` keeps everything in memory. Normally `adept()` also writes a
two-sheet Excel workbook (all segments, and a summary of the confirmed
plateaus) and a PDF per zircon.

```{r}
result$summary[, c("Final serial number", "Integration time",
                   "Final age (Ma)",
                   "Final total uncertainty incl. decay (Ma)",
                   "Confirmed plateaus")]
```

## Reading the depth profile

`keep_profiles = TRUE` returns the smoothed series and the plateau table per
zircon, which is what you need for a figure.

```{r fig.height = 3.6}
r <- adept(input, output_path = NA, make_plots = TRUE, keep_profiles = TRUE,
           verbose = FALSE)
# the objects in $plots carry a print() method that draws the profile
r$plots[[1]]
```

Orange dots are the raw ²⁰⁶Pb/²³⁸U ages, the black line is the LOESS trend, and
each red bar is a confirmed plateau with its uncertainty band.

## The uncertainty budget

This is the part worth reading carefully. Each plateau comes with a
decomposition, because "± 0.7 Ma" on its own does not tell a reader whether
that number was measured or assumed.

```{r}
keep <- !is.na(result$summary[["Final age (Ma)"]])
result$summary[keep, c("Final age (Ma)",
                       "Random uncertainty (Ma)",
                       "Systematic uncertainty (Ma)",
                       "Decay constant uncertainty (Ma)",
                       "Relative uncertainty (%)")]
```

- **Random** is what you compare ages with inside a session: the scatter of the
  points inside the plateau, plus the reproducibility of the primary reference
  material (`calibration_uncertainty`, 3 % by default).
- **Systematic** is the decay-constant contribution. It is fully correlated
  between samples, so it does not cancel when you average ages.

On this example the reference-material term dominates: the measured
within-plateau scatter contributes only a few percent of the total. **The
default 3 % is an assumption, not a measurement** — replace it with your
laboratory's long-term reproducibility before quoting the age, and say where
the number came from.

```{r}
smaller <- adept(input, calibration_uncertainty = 0.015,
                 output_path = NA, make_plots = FALSE, verbose = FALSE)
data.frame(assumed = c("3 %", "1.5 %"),
           random = c(result$summary[["Random uncertainty (Ma)"]][1],
                      smaller$summary[["Random uncertainty (Ma)"]][1]),
           age = result$summary[["Final age (Ma)"]][1])
```

The age itself does not move; only the uncertainty does.

## Choosing the parameters, and knowing how much they matter

Two parameters control how many plateaus you get:

- `variance_threshold` (default 0.1192) — how internally consistent a segment
  has to be;
- `min_plateau_resolution` (default 5 s) — how long a plateau has to last.

Both defaults sit close to a cliff. `adept_sensitivity()` maps out the
neighbourhood so you can report it rather than discover it in review.

```{r}
s <- adept_sensitivity(input,
                       variance_threshold = c(0.08, 0.1192, 0.25),
                       min_plateau_resolution = c(5, 10),
                       filter_direction = "Forward",
                       verbose = FALSE)
s[, c("variance_threshold", "min_plateau_resolution", "n_plateaus", "ages")]
```

Put a table like this in your supplementary material. If the plateau count
changes from 0 to 4 across a plausible range of a parameter, the reader
deserves to know.

`filter_direction` is the third choice: `"Forward"` keeps ages that increase
downhole, `"Reverse"` keeps ages that decrease. It is not a cosmetic switch —
on this example Forward returns three plateaus and Reverse returns one — so it
has to be justified from the CL images, not from which answer you prefer.

## Pre-processing options

Two pipelines are available:

- `preprocess = "arima_loess"` (default) — screen outliers with an ARIMA
  residual test, then fit an ordinary LOESS.
- `preprocess = "robust_loess"` — skip the screening and fit
  `loess(family = "symmetric")`, which downweights outliers instead of deleting
  them. This removes the least stable step, since it has no model-order search
  to go wrong.

```{r}
rb <- adept(input, preprocess = "robust_loess", output_path = NA,
            make_plots = FALSE, verbose = FALSE)
data.frame(preprocess = c("arima_loess", "robust_loess"),
           plateaus = c(nrow(result$summary), nrow(rb$summary)),
           ages = c(paste(signif(result$summary[["Final age (Ma)"]], 6),
                          collapse = "; "),
                    paste(signif(rb$summary[["Final age (Ma)"]], 6),
                          collapse = "; ")))
```

Run both and report the comparison. If the geological conclusion changes
between them, the conclusion was never robust.

There is a third option, `smooth = "none"`, which turns smoothing off
altogether and segments the profile as measured. It exists for data that have
already been corrected for down-hole fractionation before they reach ADEPT -
say an F(tau) correction applied by an external reduction. Such a profile is
already flat inside a domain, and fitting LOESS to it a second time would
round off the real domain boundaries.

```{r}
ns <- adept(input, smooth = "none", output_path = NA, make_plots = FALSE,
            verbose = FALSE)
data.frame(setting = c("smooth = \"loess\" (default)", "smooth = \"none\""),
           plateaus = c(nrow(result$summary), nrow(ns$summary)),
           first_age = c(result$summary[["Final age (Ma)"]][1],
                         ns$summary[["Final age (Ma)"]][1]))
```

The ages move by a small amount rather than changing the answer. What changes
is that the segmentation follows the profile you gave it, instead of a
smoothed version of it. On uncorrected data, leave the default.

## Bayesian posterior (optional)

`mcmc = TRUE` adds a piecewise-linear changepoint posterior via the `mcp`
package (which needs JAGS). It is entirely optional: without `mcp` installed
the MCMC columns are `NA` and the rest of the pipeline runs normally.

## Before you interpret the ages

A plateau is the best piecewise-constant description of the data under the
parameters you chose. That is not the same as a geological event.

For zircons that have experienced multi-stage disturbance — uranium ore,
hydrothermally altered grains — the U-Pb system may not have behaved as a
closed system in any interval, and no amount of segmentation will recover a
meaningful age. Always check the plateaus against CL/BSE imaging and the trace
element profiles (`Th_U_Mean`, `Pb_Mean` and friends are carried through for
exactly this).

Where the package can help it does: ages outside 0–4568 Ma, non-monotonic
`Time`, and a `Time` axis that looks like milliseconds are all reported before
the model is fitted.

## References

- Wang, Y. et al. (2026). Automated Depth Profiling Technique. *JGR: Solid
  Earth*, 131, e2025JB033324.
- Jaffey, A. H. et al. (1971). Precision measurement of half-lives and specific
  activities of ²³⁵U and ²³⁸U. *Physical Review C*, 4, 1889.
- Hiess, J. et al. (2012). ²³⁸U/²³⁵U systematics in terrestrial uranium-bearing
  minerals. *Science*, 335, 1610.
- Killick, R., Fearnhead, P. & Eckley, I. A. (2012). Optimal detection of
  changepoints with a linear computational cost. *JASA*, 107, 1590.
